Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2018-01-07 docs/manuscript.tex
Updates to MS. Still need to sort out some test failure with a dependency package

%% - \Abstract{} almost as usual
\Abstract{
Research synthesis, especially, in the form of meta-analysis requires data extraction from primary studies. Meta-analysis synthesizes effect sizes, often calculated from summary statistics of studies. However, exact values of such statistics are commonly hidden in figures. The R package metaDigitise extracts descriptive statistics such as means, standard deviations and, if applicable, correlations from the four types of plots: 1) mean and error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. The package interactively guides the user through data extraction process. Notably, it enables a large-scale extraction using image files, letting the user stop processing, edit and add to the resulting data fame at any point. Further, it facilitates reproducible data extraction from plots with little inter-observer bias, thus, allowing a group of people to participate the extraction of data collaboratively.
}

@@

Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are used to present data. This means that a large amount of downstream data manipulation is subsequently required, that is different across plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata

Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput

@@

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise}, which is designed for large scale data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools specific to data extraction from common plot types (mean and error plots, box plots, scatter plots and histograms, see Figure \ref{fig:plot_type}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. 
%% need to be careful abiou th use of the term 'raw data' as internally this means uncalibrated data
It also provides users with options to conduct the necessary calculations on processed data immediately after extraction so that comparable summary statistics can be obtained quickly. \pkg{metaDigitise} condenses summary data extracted from multiple figures into a single data frame which can be can easily be exported. Processed data can also be easily extracted and analysed in any way the user desires in downstream analysis within \proglang{R}. Conveniently, when needing to process many figures at different times \pkg{metaDigitise} will only import figures not already completed within a directory. This makes it easy to add new figures at any time. \pkg{metaDigitise} has also been built for reproducibility in mind. It has functions that allow users to redraw their digitisations on figures, make corrections and access the raw calibration data which is written automatically for each figure that is digitised into a special folder within the directory. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analysts to update meta-analyses more easily.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
Research synthesis, especially in the form of meta-analysis, requires data extraction from primary studies. Meta-analysis synthesizes effect sizes, often calculated from summary statistics of studies. However, exact values of such statistics are commonly hidden in figures. The R package \pkg{metaDigitise} extracts descriptive statistics such as means, standard deviations and, if applicable, correlations from the four types of plots: 1) mean and error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. The package interactively guides the user through data extraction process. Notably, it enables a large-scale extraction using image files, letting the user stop processing, edit and add to the resulting data fame at any point. Further, it facilitates reproducible data extraction from plots with little inter-observer bias, thus, allowing a group of people to participate the extraction of data collaboratively.
}

@@

First, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are used to present data. This means that a large amount of downstream data manipulation is subsequently required, that is different across plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata

Second, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput

@@

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise}, which is designed for large scale data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools specific to data extraction from common plot types (mean and error plots, box plots, scatter plots and histograms, see Figure \ref{fig:plot_type}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. 
%% need to be careful abiou th use of the term 'raw data' as internally this means uncalibrated data
It also provides users with options to conduct the necessary calculations on processed data immediately after extraction so that comparable summary statistics can be obtained quickly. \pkg{metaDigitise} condenses summary data extracted from multiple figures into a single data frame which can be can easily exported. Processed data can also be easily extracted and analysed in any way the user desires in downstream analysis within \proglang{R}. Conveniently, when needing to process many figures at different times \pkg{metaDigitise} will only import figures not already completed within a directory. This makes it easy to add new figures at any time. \pkg{metaDigitise} has also been built for reproducibility in mind. It has functions that allow users to redraw their digitisations on figures, make corrections and access the raw calibration data which is written automatically for each figure that is digitised into a special folder within the directory. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analysts to update meta-analyses more easily.
